---
title: "Behavior - prela"
author: "Valeria Bonapersona"
date: "10/14/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
library(metafor) # for meta-analysis related things
library(knitr) # for pretty talbes in markdown
library(bain)
options(scipen=1, digits=3)
```

```{r env preparation}
source("config/utilities.R")
source("src/general_funs.R")

mab <- readRDS(paste0(temp, "mab_aggr.RDS"))

```

```{r dat freq}
# frequencies
dat_freq <- mab %>% 
  filter(sex == "male") %>% 
  group_by(trauma_presence, out_grouped) %>% 
  summarize(
    n_id = length(unique(id)), 
    n_comp = length(id), 
    n_tot = sum(n_c, n_e),
    .groups = "drop"
  ) %>% 
  mutate(group = c("A", "B", "C", "D")) %>% 
  relocate(group)

```


```{r expectations}
# find all possibilities of the difference between ELA and control for
# all the groups described by dat_freq
set <- c("+", "ns", "-") # increase, ns, decrease
set_all <- rep(list(set), 4)
all_possibilities <- expand.grid(set_all)
names(all_possibilities) <- c("A", "B", "C", "D")

# remove unnecessary
rm(set, set_all)
```

If you consider that each difference between ELA and control could result in increase, not significant and decrease, there are `r nrow(all_possibilities)` possibilities in which to find them. This is considering only their relationship against 0, and not their (in)equality constraints. 



```{r prepare blind}
# filter 
dat <- mab %>% 
  mutate(
    # blinding
    blind_me = sample(1:2, nrow(mab), replace = T),
    yi = ifelse(blind_me %% 2 == 0, yi * -1, yi)
  )

# effect estimates for each group (main analysis)
dat_aggr <- dat %>% 
  filter(sex == "male") %>%
  left_join(dat_freq, by = c("trauma_presence", "out_grouped"))

# one eff size for each group
dat_aggr <- aggregate(dat_aggr, cluster = group, struct = "ID") # clustering based on inverse variance > gives same result as rma
## of note: due to the aggregation, most descriptive vars do not have any meaning at this stage

# only info of interest
dat_aggr <- dat_aggr %>% 
  select(group, out_grouped, trauma_presence, n_id, n_comp, n_tot, yi, vi) %>% 
  arrange(group) %>%
  
  # important vars for bain
  mutate(
    sampm = n_tot,
    varm = vi, # CHECK
    covm = varm/sampm
  )

# get cov matrices (1x1)
covm <- list()
for(i in 1:nrow(dat_aggr)) {
  covm[[i]] <- as.matrix(dat_aggr$covm[[i]])
}

## save estimates in a separate var
#estm <- abs(dat_aggr$yi) # make all absolute to remove from hypothesis
estm <- dat_aggr$yi
names(estm) <- dat_aggr$group

```

## Blinding
```{r bain blind}
# run bain
set.seed(100)

# write hypothesis as bain wants it
my_hyp <- paste0(rep(c("A", "B", "C", "D"), each = 2), 
                 rep(c(">", "=", "<"),4), 0, 
                 collapse = "; ")

# run bain
res_blind <- bain(estm, my_hyp, 
                n=dat_aggr$sampm,
                Sigma=covm,
                group_parameters=1,joint_parameters = 0)
```

From the blinding, A, B, C, D all came out positive. I therefore select those hypotheses that are still possible given this condition. 

There are many hypotheses to test, and some of them need to be preprocessed differently, and therefore cannot be tested together. I therefore decided to use a two step approach. First, I look at the direction of the effects. If the effects are in the same / opposite direction, the relevant hypotheses will be different. 


```{r prepare not blind}
# filter 
dat <- mab %>% 
  mutate(
    # blinding
    # blind_me = sample(1:2, nrow(mab), replace = T),
    # yi = ifelse(blind_me %% 2 == 0, yi * -1, yi)
  )

# effect estimates for each group (main analysis)
dat_aggr <- dat %>% 
  filter(sex == "male") %>%
  left_join(dat_freq, by = c("trauma_presence", "out_grouped"))

# one eff size for each group
dat_aggr <- aggregate(dat_aggr, cluster = group, struct = "ID") # clustering based on inverse variance > gives same result as rma
## of note: due to the aggregation, most descriptive vars do not have any meaning at this stage

# only info of interest
dat_aggr <- dat_aggr %>% 
  select(group, out_grouped, trauma_presence, n_id, n_comp, n_tot, yi, vi) %>% 
  arrange(group) %>%
  
  # important vars for bain
  mutate(
    sampm = n_tot,
    varm = vi, # CHECK
    covm = varm/sampm
  )

# get cov matrices (1x1)
covm <- list()
for(i in 1:nrow(dat_aggr)) {
  covm[[i]] <- as.matrix(dat_aggr$covm[[i]])
}

## save estimates in a separate var
#estm <- abs(dat_aggr$yi) # make all absolute to remove from hypothesis
estm <- dat_aggr$yi
names(estm) <- dat_aggr$group

```

## Decision tree approach
I made a decision tree scheme. Each node is a decision (aka informative hypothesis) that can be only yes or no. Going down the tree, you establish which hypothesis / group of hypotheses are most likely. I then intend to use the other outcomes to establish which one is most likely. 

```{r bain tree}
# run bain
set.seed(100)
min_eff <- 0.2
# hyp 1: abs(A) > 0
hyp_1 <- paste0("A > ", min_eff) # preprocess with abs

# run bain
res_1 <- bain(abs(estm), hyp_1, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)

## res: abs(A) > 0
# hyp 2: abs(C) > 0
hyp_2 <- paste0("C > ", min_eff)
res_2 <- bain(abs(estm), hyp_2, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)

## res 2: abs(C) > 0
# hyp 3: A and C same direction
hyp_3 <- c("A < 0 & C < 0; 
            A < 0 & C > 0;
            A > 0 & C < 0;
            A > 0 & C > 0")
res_3 <- bain(estm, hyp_3, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)
## res A < 0 & C < 0

# hyp 4: abs(A) is >, = or < than abs(C)
hyp_4 <- c("A > 0;
            A < 0") # A = 0 is in Hu
res_4 <- bain(abs(estm), hyp_4, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)

## res 4: abs(A) > abs(C)
# hyp 5 abs(B) > or = 0
hyp_5 <- paste0("B > ", min_eff)
res_5 <- bain(abs(estm), hyp_5, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)

# res 5: abs(B) < 0.2 (min eff)
# hyp 6 abs(D) > or = 0
hyp_6 <- paste0("D > ", min_eff)
res_6 <- bain(abs(estm), hyp_6, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)
# res 6 abs(D) > 0

# hyp 7: A and D are in the same or opposite direction
hyp_7 <- c("A < 0 & D < 0; 
            A < 0 & D > 0;
            A > 0 & D < 0;
            A > 0 & D > 0")
res_7 <- bain(estm, hyp_7, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)
# res 7: A and D are in opposite directions

## The results support a mismatch on sinsetization as preferred theory
```


## Features approach
```{r bain features}
## match - mismatch
# run bain
set.seed(100)
min_eff <- 0.2
# the first 4 give evidence to match-mismatch on sensitization, the second 4 on match-mismatch on hit
hyp_dir <- c("A > 0 & B < 0;
              A < 0 & B > 0;
              C > 0 & D < 0;
              C < 0 & D > 0;
             
              A > 0 & C < 0;
              A < 0 & C > 0;
              B > 0 & D < 0;
              B < 0 & D > 0;
             ")

# run bain
res_dir <- bain(estm, hyp_dir, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)

## these results show that there is evidence of match mismatch on sensitization

## hit and sens
### next I want to ask if other theories may be compatible with this theory
### only ELA and sensitization are not compatible
### multiple hits is in the first 3 hyp
### already stressed not compatible
### ceiling not compatible
hyp_comp <- c("A < C & B < D;
              A < C;
              A < D")
res_comp <- bain(abs(estm), hyp_comp, 
              n=dat_aggr$sampm,
              Sigma=covm,
              group_parameters=1,joint_parameters = 0)
### multiply PMPb together?
## from this I conclude that it is unlikely for multiple hits to increase the effects

```

With two different methods I reached the same conclusion: the match/mismatch hypothesis is the most likely. 

Although above I discussed the match / mismatch in terms of sensitization (effects are opposite for neutral vs stressful learning), I think that this is not correct. Both the neutral and stressful learning tasks are "activated", because the animal *is* doing the task. 

Perhaps this is the first descrepancy that we see between the behavior and explaining the underlying biology: the majority of bio data we have is at rest, or can be considered "activated" depending of how the animals were killed (e.g. first vs last in a cage).





## Below here it did not work: direct theories
Limitations: not all theories can be assessed at the same time "equally". Basically no information comes out if the theories are written wrong. Not informative. 



```{r imp theories to hyp}
theories <- read.csv(paste0(raw, "theories.csv"))

theories <- theories %>% 
  mutate(
    hyp_n = 1:nrow(.)
  ) %>% 
  filter(hyp_n %in% c(2:4,10:16)) # 9 and 2 are the same
```

I feel that the way I have built my hypotheses is too hard, and in this way they cannot be tested by bain. 


```{r bain}
# run bain
set.seed(100)

#min_g <- 0.2 # min effect size

# write hypothesis as bain wants it
my_hyp <- paste(theories$hyp, collapse = "; ")

# run bain
res_two <- bain(estm, my_hyp, 
                n=dat_aggr$sampm,
                Sigma=covm,
                group_parameters=1,joint_parameters = 0)
```




```{r perform meta}
# dat <- mab %>% 
#   mutate(
#     # put eff sizes in correct direction
#     # multiply = ifelse(outcome == "fear_conditioning", 1, -1),
#     # yi = yi*multiply, 
#     # blinding
#     yi = ifelse(as.numeric(id) + sample(c(1,2), size =1) %% 2 == 0, yi * -1, yi) 
#   )

mod <- rma.mv(
  yi, vi,
  mods = ~out_grouped:trauma_presence - 1,
  data = dat %>% filter(sex == "male"), 
)
```



